### 2021-04-30 doc/VerticalArrangements.tex
added additional updates

We focus first on the results for consumers in the left-most panel of Figure \ref{fig:surplussum}.  The median change is negative, indicating harm, across all four types of mergers for both cost specifications.  However, the distributions and magnitudes differ.  In particular, there is only a partial rank-ordering of consumer harm across different types of mergers: consumer harm from integrated mergers first-order stochastically dominates consumer harm from all other merger  types, while downstream and upstream mergers each first-order stochastically dominate consumer harm from vertical mergers, but do not stochastically dominate one other.   Median consumer harm from integrated mergers is about 16\% of pre-merger total expenditures, 1.37 times the magnitude of that from downstream mergers, 2.4 times the magnitude of that from upstream mergers, and more than 3.5 times the magnitude of vertical mergers.

\textbf{Is it worth exploring which effect dominates: the lessening of EDM or the strengthening of RRC? Maybe do this by allowing assymetric cost structures between merging and non-merging parties?}

@@

Turning to retailers in the second panel of Figure \ref{fig:surplussum}, we find that while downstream mergers, vertical mergers and integrated mergers always benefit retailers, upstream mergers  harm retailers in about 70\% of all simulations. Moreover, there is a partial rank-ordering across mergers that is distinct from the consumer rank-ordering: the retailer surplus distribution from integrated mergers first-order stochastically dominates the retailer surplus distribution from downstream mergers, which dominates the retailer surplus distribution from upstream mergers, but not the retailer surplus distribution from vertical mergers.

@@

As for wholesaler surplus, which appears in the third panel, the effects seen there are largely the reflection of those for retailers: wholesaler surplus increases in about 82\% of all upstream simulated mergers, 28\% of simulated vertical mergers, 20\% of simulated downstream mergers, and 17\% of simulated integrated mergers. %\textbf{comparing these results to \citeN{ST2020}, the effects are less stark (i.e. less of a transfer.}
However, only wholesaler surplus from upstream mergers first-order stochastically dominates the wholesaler surplus of the other merger types.

@@

In terms of total welfare, with the exception of the approximately 10\% of vertical mergers and 8.5\% of integrated mergers with constant costs that are beneficial, our simulated mergers are almost always net harmful. Moreover, there is a partial rank ordering of mergers, with total harm from downstream mergers first-order stochastically dominating total harm from upstream mergers, which dominates totals harm from vertical mergers. Median consumer harm from integrated mergers is about 10\% of pre-merger total expenditures, X.XX times the magnitude of that from downstream mergers, X.X times the magnitude of that from upstream mergers, and more than X.X times the magnitude of vertical mergers.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Vertically Integrated Incumbent Firms}
Here, we examine merger outcomes when a vertically integrated merges with either an unintegrated upstream or unintegrated downstream firm, or another integrated firm. We also examine merger outcomes for  mergers involving at least one integrated firm as the number of integrated non-merging rivals increases.

Figure \ref{fig:CVvertincumbBW_updownBW} depicts box and whisker plots summarizing consumer harm (top) and total harm (bottom) as the number of incumbent integrated firms increases from 0 to 6 firms. The box and whisker  plots when the number of  incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2020}.\footnote{Important differences between the simulations depicted in Figure \ref{fig:CVvertincumbBW} and those in Figure 1 of \citeN{ST2020} is that these results exclude downstream markets where prices are set according to a 2nd score auction}  For vertical mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when a single 3rd party is integrated. By contrast,for downstream and upstream  mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, with no integrated third parties. Finally, for integrated mergers, the plots when the number of  incumbent integrated firms equals 2 depict a merger between two integrated firms, again with no integrated rivals.

Starting with vertical mergers, absent incumbent integration (0 integrated firms), about 22\% of vertical mergers benefit consumers and about 14\% of vertical mergers increase total surplus.  The presence of a single 3rd party integrated firm (1 integrated firm) decreases the percentage of mergers that benefit consumers to 14.5\% and the percentage of mergers that increase  total surplus to 9.6\%, but leads median harm largely unchanged. Interestingly, while adding a second integrated incumbent to the market also lowers the percentage of beneficial mergers,adding a third integrated firm increases the percentage of mergers that benefits consumers to about 15\%, and the percentage of mergers that increase total surplus to 6.9\%. Adding a fourth integrated incumbent has a similar effect to adding a third incumbent. Taken together, Figure \ref{fig:CVvertincumbBW_updownBW} indicates that the presence of incumbent integrated firms is harmful, and that the harm does not lessen as additional additional integrated incumbents are added.

By contrast, not only are upstream mergers absent incumbent vertical integration rarely beneficial, but median consumer harm  from upstream mergers is about 6\% of pre-merger revenues  and total harm is about 3\% of pre-merger revenues. A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the proportion of beneficial mergers but does little to change median harm: consumer welfare is positive in less than 8\% of mergers and total welfare is positive in 2.3\% of mergers, but median consumer harm increases to 7\% and median total harm increases to 4.8\% of pre-merger revenues. Increasing the number of integrated 3rd parties to 1 (2 integrated firms) shrinks the inter-quartile range but has little impact on median harm. Adding additional integrated 3rd parties (3,4 integrated firms) further decreases the range of outcomes while leaving median harm largely unchanged. In summary, additional incumbent integrated firms do not substantially impact median harm, but do substantially reduce the likelihood of observing extremely harmful mergers.

Like upstream mergers, downstream mergers absent incumbent vertical integration are rarely beneficial, with median consumer harm from downstream mergers equal to about 9\% of pre-merger revenues while total harm is about 5\% of pre-merger revenues. A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the range of outcomes while also increasing median harm: consumer welfare is positive in more than 13\% of mergers and total welfare is positive in 3.4\%, but median consumer harm increases to 14\% and median total harm increases to 7.6\% of pre-merger revenues. Increasing the number of integrated 3rd parties to 1 (2 integrated firms) shrinks the inter-quartile range but has little impact on median harm. Adding additional integrated 3rd parties (3 integrated firms) further decreases the range of outcomes while also decreasing median harm. With 3 integrated 3rd parties (4 integrated firms), a downstream merger between an integrated and unintegrated firms benefits consumers in about 20\% of markets and shrinks median consumer harm to about 6\% of pre-merger revenues. By contrast, with 3 integrated 3rd parties, the inter-quartile range shrinks, and median total harm falls to about 3.7\% of pre-merger revenues.

Finally, absent the presence of rival incumbent integrated firms, mergers between two integrated firms harm consumers in about 96\% of simulated markets and lowers total surplus in more than 92\% of simulated markets. While adding one additional rival incumbent integrated firm (3 integrated firms) has little effect on either the distribution of consumer or total surplus, additional integrated rivals eliminate the most harmful integrated mergers while also increasing median consumer harm by as much as 13.3\% (5 integrated firms) and median total harm by as much as 8.4\%.

@@

Figure \ref{fig:CVbargnobothBW}  depicts box and whisker plots summarizing the consumer (top) and total (bottom) welfare effects for downstream, upstream, and vertical mergers as the number of incumbent integrated firms increases from 0 to 1 and finally 4 incumbent vertically integrated firms.  Plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated: these are comparable to those in \citeN{ST2020}. For downstream (left) and upstream (middle) mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, and all 3rd parties are unintegrated. By contrast, for vertical mergers (right), the plots when the number of  incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when a single 3rd party is integrated. Finally, plots with 4 integrated firms increases the number of rival incumbent integrated firms in upstream and downstream mergers to 3 and in vertical mergers to 4 .

@@

Turning to upstream mergers, when no incumbent firms are in the market (0 integrated firms), there is a negative relationship between bargaining power and both consumer and total harm.  Median consumer harm increases from about 2\% of pre-merger expenditures when retailers have relatively more bargaining power to about 31\% of pre-merger expenditures when wholesalers have relatively more bargaining power. Likewise, median total harm increases from 0.7\% of pre-merger expenditures when retailers have relatively more bargaining power to 14\% when wholesalers have relatively more bargaining power. By contrast, for mergers between an integrated and unintegrated upstream supplier (1 integrated firms), there is a positive relationship between bargaining power and both consumer and total surplus. Median consumer harm increases from about 8.5\% of pre-merger expenditures when retailers have relatively more bargaining power to about -20\% of pre-merger expenditures when wholesalers have relatively more bargaining power. Likewise, median total harm increases from about 3.5\% of pre-merger expenditures when retailers have relatively more bargaining power to -4.4\% when wholesalers have relatively more bargaining power. Adding 2 additional incumbent integrated firms (4 integrated firms only strengthens the positive relationship between bargaining power and harm.

Unlike downstream or upstream mergers, where the presence of vertically integrated incumbent firms can affect either the magnitude or direction of the relationship between bargaining power and harm, the presence of integrated incumbents has little impact on the strong, positive relationship between bargaining power and harm for vertical mergers. Figure \ref{fig:CVbargnobothBW} shows that consumer harm when there are fewer incumbent integrated firms first-order stochastically dominates consumer harm when there are more incumbent integrated firms, the difference is smallest when wholesalers and retailers have equal bargaining power and greatest as wholesaler power increases relative to retailer power.

>>>>

We focus first on the results for consumers in the left-most panel of Figure \ref{fig:surplussum}.  The median change is negative, indicating harm, across all four types of mergers for both cost specifications.  However, the distributions and magnitudes differ.  In particular, there is only a partial rank-ordering of consumer harm across different types of mergers: consumer harm from integrated mergers first-order stochastically dominates consumer harm from all other merger  types, while downstream and upstream mergers each first-order stochastically dominate consumer harm from vertical mergers, but do not stochastically dominate one other.   Median consumer harm from integrated mergers is almost 14\% of pre-merger total expenditures, 1.25 times the magnitude of that from downstream mergers, 2.3 times the magnitude of that from upstream mergers, and more than 4 times the magnitude of vertical mergers.

%\textbf{Is it worth exploring which effect dominates: the lessening of EDM or the strengthening of RRC? Maybe do this by allowing assymetric cost structures between merging and non-merging parties?}

@@

Turning to retailers in the second panel of Figure \ref{fig:surplussum}, we find that while downstream mergers, vertical mergers and integrated mergers always benefit retailers, upstream mergers  harm retailers in about 67\% of all simulations. Moreover, there is a partial rank-ordering across mergers that is distinct from the consumer rank-ordering: the retailer surplus distribution from integrated mergers first-order stochastically dominates the retailer surplus distribution from downstream mergers, which dominates the retailer surplus distribution from upstream mergers, but not the retailer surplus distribution from vertical mergers.

@@

As for wholesaler surplus, which appears in the third panel, the effects seen there are largely the reflection of those for retailers: wholesaler surplus increases in about 77\% of all upstream simulated mergers, 23\% of simulated vertical mergers, 20\% of simulated downstream mergers, and abput 15\% of simulated integrated mergers. %\textbf{comparing these results to \citeN{ST2020}, the effects are less stark (i.e. less of a transfer.}
However, only wholesaler surplus from upstream mergers first-order stochastically dominates the wholesaler surplus of the other merger types.

@@

In terms of total welfare, approximately 18\% of vertical mergers and 15\% of integrated mergers with constant costs that are beneficial, whereas only 3\% of upstream mergers and 2\% of downstream mergers are beneficial. Moreover, there is a partial rank ordering of mergers, with total harm from downstream mergers first-order stochastically dominating total harm from upstream mergers, which dominates totals harm from vertical mergers. Median consumer harm from integrated mergers is about 9.6\% of pre-merger total expenditures, 1.5 times the magnitude of that from downstream mergers, 2.5 times the magnitude of that from upstream mergers, and more than 4.3 times the magnitude of vertical mergers.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Vertically Integrated Incumbent Firms}
Here, we examine merger outcomes when a vertically integrated merges with either an unintegrated upstream or unintegrated downstream firm, or another integrated firm. We also examine merger outcomes for  mergers as the number of incumbent integrated non-merging rivals increases.

Figure \ref{fig:CVvertincumbBW_updownBW} depicts box and whisker plots summarizing consumer harm (top) and total harm (bottom) as the number of incumbent integrated firms increases from 0 to 6 firms. The box and whisker  plots when the number of  incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2020} and is included as reference.\footnote{An important difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW} and those in Figure 1 of \citeN{ST2020} is that these results exclude downstream markets where prices are set according to a 2nd score auction}  For vertical mergers, the plots when the number of  incumbent integrated firms equals 1 summarize market outcomes where an unintegrated wholesaler merging with an unintegrated retailer, a single unintegrated 3rd party is integrated, and the remaining 3rd parties (if any) are unintegrated. By contrast,for downstream and upstream  mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, with no integrated third parties. Finally, for integrated mergers, the plots when the number of  incumbent integrated firms equals 2 depict a merger between two integrated firms, again with no integrated rivals.

Starting with vertical mergers, absent incumbent integration (0 integrated firms), about 36\% of vertical mergers benefit consumers and about 23\% of vertical mergers are net beneficial (i.e. increase total surplus).  The presence of a single 3rd party integrated firm (1 integrated firm) decreases the percentage of mergers that benefit consumers to 25\% and the percentage of mergers that increase  total surplus to 16\%, but leads median harm largely unchanged. Interestingly, while adding a second integrated incumbent to the market also lowers the percentage of beneficial mergers, adding a third integrated firm increases the percentage of mergers that benefits consumers to about 27\% but does not increase the percentage of net beneficial mergers. Adding a fourth integrated incumbent increases the percentage of mergers that benefit consumers to 32\% but decreases the percentage of netbeneficial mergers to about 11\%. Taken together, Figure \ref{fig:CVvertincumbBW_updownBW} indicates that the presence of incumbent integrated firms is harmful, and that the harm does not meaningfully lessen as additional integrated incumbents are included.

By contrast, not only are upstream mergers absent incumbent vertical integration rarely beneficial, but median consumer harm  from upstream mergers is about 6\% of pre-merger revenues  and total harm is about 3\% of pre-merger revenues. A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the proportion of beneficial mergers but does little to change median harm: consumer welfare increases in almost 16\% of mergers and total welfare increases in 4.3\% of mergers, but median consumer harm increases to 6.3\% and median total harm increases to 4.8\% of pre-merger revenues. Increasing the number of integrated 3rd parties to 1 (2 integrated firms) disproportionately shrinks the inter-quartile range-- relatively more harmful mergers are eliminated than beneficial ones-- but the presence of a single integrated 3rd party has little impact on median harm. Adding additional integrated 3rd parties (3,4 integrated firms) further decreases the range of outcomes while leaving median harm largely unchanged. In summary, additional incumbent integrated firms do not substantially impact median harm, but do substantially reduce the likelihood of observing extremely harmful mergers.

Like upstream mergers, downstream mergers absent incumbent vertical integration are rarely beneficial, with median consumer harm from downstream mergers equal to about 9\% of pre-merger revenues while total harm is about 5\% of pre-merger revenues. A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the range of outcomes while also increasing median harm: consumer welfare increases in about 18\% of mergers and total welfare increases in 5.7\%, but median consumer harm increases to 14\% and median total harm increases to 7.3\% of pre-merger revenues. Increasing the number of integrated 3rd parties to 1 (2 integrated firms) shrinks the inter-quartile range but has little impact on median harm. Adding additional integrated 3rd parties (3 integrated firms) further decreases the range of outcomes while also decreasing median harm t 5.6\%. With 3 integrated 3rd parties (4 integrated firms), a downstream merger between an integrated and unintegrated firms benefits consumers in about 27\% of markets and shrinks median consumer harm to 6.4\% of pre-merger revenues. By contrast, with 3 integrated 3rd parties, the inter-quartile range of total harm shrinks, and median total harm falls to about 3.8\% of pre-merger revenues. These results indicate that unlike either vertical upstream mergers, integrated downstream mergers are more harmful than unintegrated downstream mergers, and that the presence of additional integrated rival incumbents reduces harm, though slowly.

Finally, absent the presence of rival incumbent integrated firms, mergers between two integrated firms benefits consumers in about 7\% of simulated markets and is net beneficial in 15\% of simulated markets. While adding one additional rival incumbent integrated firm (3 integrated firms) has little effect on either the distribution of either outcome, adding a second integrated rival (4 integrated firms) eliminates the most harmful integrated mergers while also increasing median consumer harm to 14\%  and median total harm to 10.6\%. The presence of additional integrated incumbents has little incremental effect on  median harm.

@@

Figure \ref{fig:CVbargnobothBW}  depicts box and whisker plots summarizing the consumer (top) and total (bottom) welfare effects for downstream, upstream, and vertical mergers as the relative bargaining power parameter increases from 1/9 (retailers have the advantage to 9 (wholesalers have the advantage). also depicted are three sets of box and whisker plots that correspond to the number of incumbent integrated firms included in the simulated markets: 0 (light blue), 1 (darker blue), and 4 (darkest blue)incumbent vertically integrated firms.  As in Figure \ref{fig:CVvertincumbBW_updownBW}, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated: these are comparable to those in \citeN{ST2020}. For vertical mergers (left), the plots when the number of  incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when a single 3rd party is integrated. By contrast, for upstream (middle) and downstream (left) mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, and all 3rd parties are unintegrated. Finally, plots with 4 integrated firms increases the number of rival incumbent integrated firms in upstream and downstream mergers to 3 and in vertical mergers to 4 .

For vertical mergers, the presence of integrated incumbents has little impact on the strong, positive relationship between bargaining power and harm for vertical mergers. Figure \ref{fig:CVbargnobothBW} shows that consumer harm when there are fewer incumbent integrated firms first-order stochastically dominates consumer harm when there are more incumbent integrated firms, the difference is smallest when wholesalers and retailers have equal bargaining power and greatest as wholesaler power increases relative to retailer power. Across all three incumbent integrated firm scenarios, median consumer harm decreases from about 7\% when retailers have the advantage to -24\% when wholesalers have the advantage. Similarly, median net harm decreases from about 2.9\% of pre-merger expenditures when retailers have relatively more bargaining power to about -14\% of pre-merger expenditures when wholesalers have relatively more bargaining power.

Turning to upstream mergers, when no incumbent firms are in the market (0 integrated firms), there is a negative relationship between bargaining power and both consumer and total harm.  Median consumer harm decreases from 2.7\% of pre-merger expenditures when retailers have relatively more bargaining power to about 31\% of pre-merger expenditures when wholesalers have relatively more bargaining power. Likewise, median total harm increases from 0.7\% of pre-merger expenditures when retailers have relatively more bargaining power to 14\% when wholesalers have relatively more bargaining power. By contrast, for mergers between an integrated and unintegrated upstream supplier (1 integrated firms), there is a positive relationship between bargaining power and both consumer and total surplus. Median consumer harm decreases from about 8.5\% of pre-merger expenditures when retailers have relatively more bargaining power to about -20\% of pre-merger expenditures when wholesalers have relatively more bargaining power. Likewise, median total harm increases from about 3.5\% of pre-merger expenditures when retailers have relatively more bargaining power to -4.4\% when wholesalers have relatively more bargaining power. Adding 2 additional incumbent integrated firms (4 integrated firms only strengthens the positive relationship between bargaining power and harm.

@@
